---
title: "Macro Signals and U.S. Equity Returns"
subtitle: "A release-aware, out-of-sample classification study"
author: "Som Tripathi"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    self_contained: true
    css: styles.css
    highlight: tango
params:
  analysis_start: "2000-01-01"
  train_end: "2015-12-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5,
  fig.retina = 2,
  out.width = "100%"
)

library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(tibble)
library(tidyr)
library(tidyquant)

portfolio_colors <- c(
  navy = "#12263A",
  blue = "#2563EB",
  teal = "#0F766E",
  gold = "#D97706",
  red = "#B91C1C",
  slate = "#64748B",
  pale = "#E2E8F0"
)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", color = portfolio_colors[["navy"]]),
      plot.subtitle = element_text(color = portfolio_colors[["slate"]]),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      plot.caption = element_text(color = portfolio_colors[["slate"]])
    )
)

options(scipen = 999)
source("R/analysis.R", local = knitr::knit_global())
```

<div class="hero">
<p class="eyebrow">PORTFOLIO CASE STUDY</p>
<p class="lede">Can the unemployment rate, combined with recent market behavior, improve forecasts of next-month S&amp;P 500 direction?</p>
</div>

## Executive summary

```{r executive-summary, echo=FALSE, results='asis'}
model_accuracy <- metric_value("Logistic regression", "Accuracy")
baseline_accuracy <- metric_value("Historical up-rate baseline", "Accuracy")
model_auc <- metric_value("Logistic regression", "ROC AUC")

cat(
  '<div class="metric-grid">',
  sprintf(
    '<div class="metric-card"><span class="metric-value">%s</span><span class="metric-label">Out-of-sample accuracy</span></div>',
    percent(model_accuracy, accuracy = 0.1)
  ),
  sprintf(
    '<div class="metric-card"><span class="metric-value">%s</span><span class="metric-label">Baseline accuracy</span></div>',
    percent(baseline_accuracy, accuracy = 0.1)
  ),
  sprintf(
    '<div class="metric-card"><span class="metric-value">%.2f</span><span class="metric-label">Out-of-sample ROC AUC</span></div>',
    model_auc
  ),
  sprintf(
    '<div class="metric-card"><span class="metric-value">%s</span><span class="metric-label">Test observations</span></div>',
    comma(nrow(test_data))
  ),
  '</div>'
)
```

This study uses monthly observations from **`r format(min(model_data$month_date), "%B %Y")`**
through **`r format(max(model_data$month_date), "%B %Y")`**. A logistic regression is
trained through **`r format(as.Date(params$train_end), "%B %Y")`** and evaluated only on
later observations.

> **Bottom line.** `r performance_summary` The result is useful precisely because the
> benchmark is difficult to beat: U.S. equity months are positive more often than they
> are negative, so headline accuracy without a baseline can be misleading.

## Research design

The unit of analysis is one calendar month. At the close of month \(t\), the model uses:

- the latest unemployment rate that would have been published by then;
- the latest published monthly change in unemployment;
- the S&P 500 return during month \(t\); and
- annualized daily volatility during month \(t\).

It predicts whether the S&P 500 return in month \(t+1\) will be positive.

```{r design-table, echo=FALSE}
design_table <- tribble(
  ~Component, ~Choice, ~Reason,
  "Target", "Next-month S&P 500 direction", "Clear binary outcome with an investable forecast horizon",
  "Macro timing", "Unemployment lagged one month", "Avoids using a value before its public release",
  "Validation", "Chronological holdout", "Preserves ordering and prevents random-split leakage",
  "Baseline", "Training-period up rate", "Measures whether the model beats the market's class imbalance",
  "Metrics", "Accuracy, balanced accuracy, Brier score, ROC AUC", "Covers classification and probability quality"
)

knitr::kable(design_table, align = c("l", "l", "l"))
```

### Why publication timing matters

The unemployment rate for a given month is generally released during the following
month. Treating month \(t\)'s unemployment rate as known during month \(t\) creates
look-ahead bias. This analysis conservatively maps each unemployment observation to the
next calendar month, when it is available to a forecaster.

## Data and preparation

The repository includes versioned raw-data snapshots from two public sources:

- [Yahoo Finance](https://finance.yahoo.com/quote/%5EGSPC/history) for daily adjusted
  S&P 500 prices (`^GSPC`);
- [FRED](https://fred.stlouisfed.org/series/UNRATE) for the monthly U.S. unemployment
  rate (`UNRATE`).

By default, the report renders from those snapshots so the analysis remains available
when an upstream service is unavailable. Setting `REFRESH_DATA=true` attempts a live
refresh and updates the snapshots when successful. This render used
**`r sp500_data_source`** and **`r unemployment_data_source`**.

The current, incomplete market month is excluded. Monthly returns are compounded from
daily log returns, while volatility is the standard deviation of daily log returns
annualized with \(\sqrt{252}\).

```{r coverage-table, echo=FALSE}
knitr::kable(coverage_table, align = c("l", "l", "l", "r"))
```

```{r data-preparation}
sp500_daily <- sp500_raw %>%
  arrange(date) %>%
  mutate(daily_log_return = log(adjusted / lag(adjusted))) %>%
  filter(!is.na(daily_log_return))

sp500_monthly <- sp500_daily %>%
  mutate(month_date = floor_date(date, "month")) %>%
  group_by(month_date) %>%
  summarise(
    monthly_log_return = sum(daily_log_return),
    monthly_return = exp(monthly_log_return) - 1,
    annualized_volatility = sd(daily_log_return) * sqrt(252),
    n_trading_days = n(),
    .groups = "drop"
  )
```

### Market history

```{r market-history, echo=FALSE}
ggplot(sp500_raw, aes(date, adjusted)) +
  geom_line(linewidth = 0.55, color = portfolio_colors[["blue"]]) +
  scale_y_log10(labels = label_dollar()) +
  labs(
    title = "The S&P 500 across multiple market cycles",
    subtitle = "Adjusted close shown on a logarithmic scale",
    x = NULL,
    y = "Adjusted close (log scale)",
    caption = "Source: Yahoo Finance via tidyquant"
  )
```

```{r unemployment-history, echo=FALSE}
ggplot(unemp_monthly, aes(date, unemployment_rate)) +
  geom_area(fill = portfolio_colors[["pale"]], color = portfolio_colors[["teal"]],
            linewidth = 0.55) +
  labs(
    title = "U.S. unemployment captures distinct macro regimes",
    subtitle = "The pandemic spike is retained rather than winsorized",
    x = NULL,
    y = "Unemployment rate (%)",
    caption = "Source: FRED series UNRATE via tidyquant"
  )
```

### Descriptive relationships

These plots are descriptive, not causal. They show the contemporaneous relationship
between unemployment conditions and market behavior without claiming that unemployment
alone drives returns.

```{r descriptive-relationships, echo=FALSE}
ggplot(descriptive_data, aes(unemployment_rate, monthly_return)) +
  geom_hline(yintercept = 0, color = portfolio_colors[["pale"]]) +
  geom_point(alpha = 0.55, color = portfolio_colors[["navy"]]) +
  geom_smooth(method = "lm", se = TRUE, color = portfolio_colors[["gold"]],
              fill = alpha(portfolio_colors[["gold"]], 0.15)) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    title = "Monthly returns remain noisy across unemployment levels",
    x = "Unemployment rate (%)",
    y = "S&P 500 monthly return",
    caption = "Each point represents one month"
  )
```

```{r regime-summary, echo=FALSE}
ggplot(regime_summary, aes(unemp_regime, avg_volatility, fill = unemp_regime)) +
  geom_col(width = 0.62, show.legend = FALSE) +
  geom_text(
    aes(label = percent(avg_volatility, accuracy = 0.1)),
    vjust = -0.5,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c(
    "Lower unemployment" = portfolio_colors[["blue"]],
    "Higher unemployment" = portfolio_colors[["gold"]]
  )) +
  scale_y_continuous(labels = label_percent(), expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Volatility is higher in high-unemployment regimes",
    subtitle = "Regimes use the median unemployment rate over the analysis sample",
    x = NULL,
    y = "Average annualized volatility"
  )
```

## Modeling

### Feature engineering and split

All transformations are calculated in chronological order. Predictor standardization is
fit on the training sample and then applied unchanged to the test sample.

```{r model-specification}
feature_names <- c(
  "available_unemployment",
  "available_unemp_change",
  "monthly_return",
  "annualized_volatility"
)

train_data <- model_data %>%
  filter(month_date <= floor_date(as.Date(params$train_end), "month"))

test_data <- model_data %>%
  filter(month_date > floor_date(as.Date(params$train_end), "month"))

feature_center <- vapply(train_data[feature_names], mean, numeric(1))
feature_scale <- vapply(train_data[feature_names], sd, numeric(1))

train_model <- standardize_features(train_data, feature_center, feature_scale)
test_model <- standardize_features(test_data, feature_center, feature_scale)

logit_fit <- glm(
  next_month_up ~ available_unemployment + available_unemp_change +
    monthly_return + annualized_volatility,
  data = train_model,
  family = binomial()
)
```

### Coefficient interpretation

Because predictors are standardized, each coefficient represents a one-standard-deviation
increase in that feature while the others are held constant. Confidence intervals that
cross an odds ratio of 1 indicate weak evidence of a stable directional relationship.

```{r coefficient-plot, echo=FALSE}
ggplot(coefficient_table, aes(odds_ratio, reorder(term_label, odds_ratio))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = portfolio_colors[["slate"]]) +
  geom_errorbarh(
    aes(xmin = conf_low, xmax = conf_high),
    height = 0.18,
    color = portfolio_colors[["blue"]]
  ) +
  geom_point(size = 3, color = portfolio_colors[["navy"]]) +
  scale_x_log10() +
  labs(
    title = "Estimated effect of each standardized feature",
    subtitle = "Odds ratios with 95% confidence intervals",
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

## Out-of-sample evaluation

The model is evaluated from **`r format(min(test_data$month_date), "%B %Y")`** through
**`r format(max(test_data$month_date), "%B %Y")`**. The baseline assigns every test
observation the positive-month probability observed in the training period.

```{r metrics-table, echo=FALSE}
display_metrics <- evaluation_metrics %>%
  filter(metric %in% c("Accuracy", "Balanced accuracy", "Brier score", "ROC AUC")) %>%
  mutate(value = if_else(metric == "Brier score", number(value, accuracy = 0.001),
                         percent(value, accuracy = 0.1))) %>%
  tidyr::pivot_wider(names_from = model, values_from = value)

knitr::kable(display_metrics, align = c("l", "r", "r"))
```

```{r accuracy-comparison, echo=FALSE}
evaluation_metrics %>%
  filter(metric %in% c("Accuracy", "Balanced accuracy", "ROC AUC")) %>%
  ggplot(aes(metric, value, fill = model)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.64) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = portfolio_colors[["slate"]]) +
  scale_fill_manual(values = c(
    "Logistic regression" = portfolio_colors[["blue"]],
    "Historical up-rate baseline" = portfolio_colors[["slate"]]
  )) +
  scale_y_continuous(labels = label_percent(), limits = c(0, 1)) +
  labs(
    title = "Performance must be judged against the baseline",
    subtitle = "The dashed line marks chance-level discrimination",
    x = NULL,
    y = "Score",
    fill = NULL
  )
```

### Confusion matrix

```{r confusion-matrix, echo=FALSE}
ggplot(confusion_matrix, aes(Predicted, Actual, fill = n)) +
  geom_tile(color = "white", linewidth = 2) +
  geom_text(aes(label = n), size = 7, fontface = "bold", color = "white") +
  scale_fill_gradient(low = portfolio_colors[["slate"]], high = portfolio_colors[["blue"]],
                      guide = "none") +
  coord_equal() +
  labs(
    title = "Out-of-sample confusion matrix",
    subtitle = "Counts at the fixed 0.50 classification threshold",
    x = "Predicted class",
    y = "Actual class"
  )
```

### Predicted probability through time

```{r probability-over-time, echo=FALSE}
ggplot(test_results, aes(month_date, predicted_probability)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = portfolio_colors[["slate"]]) +
  geom_line(color = portfolio_colors[["blue"]], linewidth = 0.7) +
  geom_point(aes(color = actual_label), size = 1.8, alpha = 0.85) +
  scale_color_manual(values = c(
    "Down month" = portfolio_colors[["red"]],
    "Up month" = portfolio_colors[["teal"]]
  )) +
  scale_y_continuous(labels = label_percent(), limits = c(0, 1)) +
  labs(
    title = "Predicted probability of an up month",
    subtitle = "Point color shows the realized next-month outcome",
    x = NULL,
    y = "Predicted probability",
    color = NULL
  )
```

## Conclusions

1. **The evaluation design matters more than the algorithm.** A chronological split,
   release-aware macro features, and an explicit baseline prevent an optimistic result.
2. **Unemployment is more informative about market conditions than short-horizon
   direction.** High-unemployment periods coincide with elevated volatility, but that
   relationship does not automatically translate into reliable next-month forecasts.
3. **A weak predictive result is still a useful research result.** The model provides a
   transparent benchmark and a foundation for testing richer real-time macro indicators,
   valuation measures, or walk-forward estimation.

## Limitations and next steps

- FRED observations are revised over time; this report uses the latest vintage rather
  than a real-time vintage database such as ALFRED.
- The model does not include transaction costs, portfolio sizing, or an investable
  backtest.
- One fixed train/test split is transparent but does not measure stability across
  multiple historical windows.
- Statistical significance is not economic significance, and this analysis should not
  be interpreted as investment advice.

A logical extension would use expanding-window validation and real-time macro vintages,
then compare logistic regression with regularized models while preserving the same
release-aware feature set.

## Reproducibility

The report was rendered with R **`r getRversion()`**. Package versions are recorded in
`renv.lock`; the full analysis pipeline lives in `R/analysis.R`.

```{r session-info}
sessionInfo()
```
