---
title: "Macro & Markets: Returns, Unemployment, and ML"
author: "Som Tripathi"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(lubridate)
library(tidyquant)
```


## Data: S&P 500 prices (Yahoo Finance)

We pull daily S&P 500 index prices (ticker `^GSPC`) from Yahoo Finance using `tidyquant`.
This keeps the report reproducible because the data is retrieved directly in code.

```{r get-sp500}
sp500_raw <- tq_get("^GSPC", from = "2000-01-01", get = "stock.prices")

# quick checks
dplyr::glimpse(sp500_raw)
dplyr::summarise(sp500_raw,
  start_date = min(date),
  end_date   = max(date),
  n_rows     = n()
)
```


## Cleaning: daily log returns

We compute daily log returns from the adjusted close price:

Return = log(adjusted_today / adjusted_yesterday)

This creates a stationary-ish series that’s easier to summarize and aggregate than raw prices.

```{r clean-sp500}
sp500_daily <- sp500_raw %>%
  arrange(date) %>%
  mutate(daily_log_return = log(adjusted / lag(adjusted))) %>%
  drop_na(daily_log_return)

# sanity checks
dplyr::summarise(sp500_daily,
  start_date = min(date),
  end_date   = max(date),
  n_rows     = n(),
  mean_return = mean(daily_log_return),
  sd_return   = sd(daily_log_return)
)
```


## Marginal summaries: daily market behavior

This section summarizes the distribution of daily returns and shows how the index level
changes over time.

```{r sp500-marginal-summary}
# numeric summaries
sp500_daily %>%
  summarise(
    n_days = n(),
    min_return = min(daily_log_return),
    q05_return = quantile(daily_log_return, 0.05),
    median_return = median(daily_log_return),
    mean_return = mean(daily_log_return),
    q95_return = quantile(daily_log_return, 0.95),
    max_return = max(daily_log_return)
  )
```

# price level over time

```{r}
sp500_raw %>%
ggplot(aes(x = date, y = adjusted)) +
geom_line() +
labs(
title = "S&P 500 adjusted close (2000+)",
x = "Date",
y = "Adjusted close"
)
```


# distribution of daily log returns

```{r}
sp500_daily %>%
ggplot(aes(x = daily_log_return)) +
geom_histogram(bins = 70) +
labs(
title = "Distribution of daily log returns (S&P 500)",
x = "Daily log return",
y = "Count"
)
```


## Data: Unemployment rate (FRED)

We pull monthly U.S. unemployment data (series `UNRATE`) from FRED using `tidyquant`.
This is measured in percent and is typically reported monthly.

```{r get-unemployment}
unemp_raw <- tq_get("UNRATE", from = "2000-01-01", get = "economic.data")

# quick checks
dplyr::glimpse(unemp_raw)
dplyr::summarise(unemp_raw,
  start_date = min(date),
  end_date   = max(date),
  n_rows     = n()
)
```


## Cleaning + marginal summary: unemployment rate

`tidyquant` stores the unemployment series value in a column called `price`.
We rename it to `unemployment_rate` to make the code easier to read.

```{r clean-unemployment}
unemp_monthly <- unemp_raw %>%
  rename(unemployment_rate = price) %>%
  arrange(date)

# numeric summary
unemp_monthly %>%
  summarise(
    n_months = n(),
    min_unemp = min(unemployment_rate),
    q05_unemp = quantile(unemployment_rate, 0.05),
    median_unemp = median(unemployment_rate),
    mean_unemp = mean(unemployment_rate),
    q95_unemp = quantile(unemployment_rate, 0.95),
    max_unemp = max(unemployment_rate)
  )
```
```{r}
unemp_monthly %>%
ggplot(aes(x = date, y = unemployment_rate)) +
geom_line() +
labs(
title = "U.S. Unemployment Rate (UNRATE, 2000+)",
x = "Date",
y = "Unemployment rate (%)"
)
```


## Monthly market features + join with unemployment

Unemployment is monthly, so we aggregate daily S&P 500 returns to monthly features:
- mean monthly log return (average of daily log returns within the month)
- monthly volatility proxy (sd of daily log returns within the month)

Then we join on the month.

```{r build-monthly-and-join}
sp500_monthly <- sp500_daily %>%
  mutate(month_date = floor_date(date, unit = "month")) %>%
  group_by(month_date) %>%
  summarise(
    mean_log_return = mean(daily_log_return),
    sd_log_return   = sd(daily_log_return),
    n_trading_days  = n(),
    .groups = "drop"
  )

macro_market_monthly <- sp500_monthly %>%
  inner_join(unemp_monthly, by = c("month_date" = "date"))
```
# quick checks
```{r}
dplyr::glimpse(macro_market_monthly)
dplyr::summarise(macro_market_monthly,
  start_month = min(month_date),
  end_month   = max(month_date),
  n_months    = n()
)
```



## Joint relationships: unemployment vs returns and volatility

These plots show how monthly market behavior changes across unemployment conditions.
Each point is one month.

```{r joint-plots, fig.width=7, fig.height=4}
# Returns vs unemployment
macro_market_monthly %>%
  ggplot(aes(x = unemployment_rate, y = mean_log_return)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Monthly mean log return vs unemployment rate",
    x = "Unemployment rate (%)",
    y = "Mean monthly log return"
  )
```
# Volatility vs unemployment
```{r}
macro_market_monthly %>%
  ggplot(aes(x = unemployment_rate, y = sd_log_return)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Monthly volatility (sd of daily returns) vs unemployment rate",
    x = "Unemployment rate (%)",
    y = "Monthly volatility (sd of daily log returns)"
  )
```


## Regimes: high vs low unemployment

To make the comparison clearer, we split months into "low" vs "high" unemployment.
Here we use the median unemployment rate as the cutoff.

```{r regimes}
cutoff <- median(macro_market_monthly$unemployment_rate)

macro_regime <- macro_market_monthly %>%
  mutate(
    unemp_regime = if_else(unemployment_rate > cutoff, "High unemployment", "Low unemployment")
  )
```
# summary table by regime
```{r}
regime_summary <- macro_regime %>%
  group_by(unemp_regime) %>%
  summarise(
    n_months = n(),
    avg_return = mean(mean_log_return),
    avg_volatility = mean(sd_log_return),
    .groups = "drop"
  )

regime_summary
```
# compare returns by regime

```{r}
macro_regime %>%
ggplot(aes(x = unemp_regime, y = mean_log_return)) +
geom_boxplot() +
labs(
title = "Monthly returns by unemployment regime",
x = "",
y = "Mean monthly log return"
)
```


# compare volatility by regime

```{r}
macro_regime %>%
ggplot(aes(x = unemp_regime, y = sd_log_return)) +
geom_boxplot() +
labs(
title = "Monthly volatility by unemployment regime",
x = "",
y = "Monthly volatility (sd of daily log returns)"
)
```


## ML setup: predict next month direction

We define a binary target:
- `next_month_up = 1` if next month's mean log return is positive, else 0.

Features use only information available in the current month:
- unemployment_rate
- change in unemployment (this month - last month)
- current month return and volatility

We split by time (train on earlier years, test on later years) to avoid leakage.

```{r ml-dataset}
ml_data <- macro_market_monthly %>%
  arrange(month_date) %>%
  mutate(
    next_mean_log_return = lead(mean_log_return),
    next_month_up = if_else(next_mean_log_return > 0, 1, 0),
    unemp_change = unemployment_rate - lag(unemployment_rate)
  ) %>%
  drop_na(next_mean_log_return, unemp_change)
```
# time split cutoff (feel free to change later)
```{r}
train_data <- ml_data %>% filter(year(month_date) <= 2015)
test_data  <- ml_data %>% filter(year(month_date) > 2015)
```
# quick checks
```{r}
c(nrow(train_data), nrow(test_data))
test_data %>% summarise(up_rate = mean(next_month_up))
```


## Model 1: Logistic regression

We use logistic regression to estimate the probability that next month is "up".
This is a strong baseline because it’s simple, interpretable, and easy to reproduce.

```{r fit-logistic}
logit_fit <- glm(
  next_month_up ~ unemployment_rate + unemp_change + mean_log_return + sd_log_return,
  data = train_data,
  family = binomial()
)

summary(logit_fit)
```


## Evaluation: test performance + baseline

We evaluate on the held-out test period (years > 2015).
We also compare against a simple baseline: always predict "up".

```{r eval-logistic}
# predicted probabilities for "up"
test_probs <- predict(logit_fit, newdata = test_data, type = "response")

# classify using 0.5 threshold
test_pred <- if_else(test_probs >= 0.5, 1, 0)

# confusion matrix
table(
  Actual = test_data$next_month_up,
  Predicted = test_pred
)

# accuracy
acc_model <- mean(test_pred == test_data$next_month_up)

# baseline: always predict up
baseline_pred <- rep(1, nrow(test_data))
acc_baseline <- mean(baseline_pred == test_data$next_month_up)

c(model_accuracy = acc_model, baseline_accuracy = acc_baseline)
```

In the held-out test period, the logistic regression model achieved about 65.5% accuracy, while a naive baseline that always predicts an “up” month achieved about 69.0% accuracy. This indicates that the simple feature set used here (unemployment level, unemployment change, current-month return, and current-month volatility) does not provide strong predictive power for next-month market direction. This result is expected in real financial data because returns are noisy and many drivers are not captured in a small macro-only feature set. The model is still useful as a transparent baseline and as a demonstration of a time-respecting ML workflow.


```{r prob-over-time, fig.width=7, fig.height=4}
test_plot_df <- test_data %>%
  mutate(pred_prob_up = test_probs)

test_plot_df %>%
  ggplot(aes(x = month_date, y = pred_prob_up)) +
  geom_line() +
  labs(
    title = "Predicted probability of an up month (test period)",
    x = "Month",
    y = "Predicted P(next month up)"
  )
```

